Finally Tagless, Partially Evaluated

### 2007-06-13 tagless/tagless-final.tex
arranging the code
small typographical corrections

It is commonly thought that to interpret a typed object language in
a typed metalanguage while preserving types is difficult and requires
GADTs or dependent types \citep{taha-tag}.  In fact, this problem
motivated much work on GADTs \cite{xi-guarded,peyton-jones-simple} and
on dependent types \cite{WalidICFP02,fogarty-concoqtion}.
%aplas-else
\begin{comment}
For a metalanguage's type system to allow the well-typed object term
|test1| but disallow the ill-typed object term |test2|, fancier types
such as GADTs or dependent types seem necessary.  
\end{comment}
Yet other type systems
have been proposed to distinguish closed terms like |test1| from open
terms like |test3|
%aplas \cite{WalidPOPL03,NanevskiICFP02,NanevskiJFP05,DaviesJACM01,nanevski-contextual},
\cite{WalidPOPL03,NanevskiJFP05,DaviesJACM01},
so that |lookup| never receives an empty environment.  We discuss these
proposals further in \S\ref{related}%
%aplas-specific
.
%aplas-else
\begin{comment}
; here we just note that many
advanced type systems have been devised to ensure statically that an
object term is well-typed and closed.
\end{comment}

@@

In sum, by Church\hyp encoding terms using ordinary functions, we
achieve a tagless evaluator for a typed object language in a
metalanguage with a simple 
%aplas
Hindley-Milner type system
%aplas \cite{hindley-principal,milner-theory}
.  In this \emph{final} rather
than \emph{initial} approach, both kinds of run-time errors in
\S\ref{tagproblem} (applying a nonfunction and evaluating an open
term) are reported at compile time. Because the new interpreter
uses no universal type or pattern matching, it never results in a
run-time error (and is in fact total).  Because this safety is obvious
not just to us but also to the metalanguage implementation, we avoid
the serious performance penalty \cite{WalidICFP02} of error checking.
\Citet{Gluck-jones-optimality} explains deeper technical reasons that
inevitably lead to these performance penalties.

@@

The term ``constructor'' functions |b|, |lam|, |app|, and so on appear
free in the encoding of an object term such as |testf1| above.  Defining
these functions differently gives rise to different interpreters, that
is, different folds on object programs.  Given the same term
representation but varying the interpreter, we can
\begin{itemize*}
    \item evaluate the term to a value in the metalanguage;
    \item measure the size or depth of the term;
    \item compile the term, with staging support such as in MetaOCaml;
    \item even partially evaluate the term, online; and
    \item transform the term to continuation\hyp passing style (CPS),
        including call-by-name (CBN) CPS, so to isolate the evaluation
        order of the object language from that of the metalanguage
%aplas
\footnote{Due to serious lack of space, 
we refer the reader to the accompanying code for this.}%
.
\end{itemize*}
We have programmed our interpreters in OCaml (and, for staging,
\cite{metaocaml}) and standard Haskell. The complete code is available
\url{\oleg{put URL!}.}
to supplement the paper. Our examples below switch between (Meta)OCaml
and Haskell even though we have implemented each example equivalently in
both metalanguages, because some of our claims are more obvious in one
metalanguage than the other.  For example, MetaOCaml provides
convenient, typed staging facilities.

@@

\begin{enumerate*}
\item We build interpreters that evaluate (\S\ref{language}),
    compile (or evaluate with staging) (\S\ref{compiler}), and partially evaluate (\S\ref{PE}) a typed higher-order object language
   in a typed metalanguage, in direct and continuation\hyp passing styles%
% aplas
%   (\S\ref{variations}).
.
\item All these interpreters use no type tags, patently never get stuck,
    and need no advanced type-system features such as GADTs, dependent types,
    or intentional type analysis.
\item The partial evaluator avoids polymorphic lift and delays binding-time
    analysis.  It bakes a type-to-type map into the interpreter
    interface to eliminate the need for GADTs and thus remain portable
    across Haskell 98 and ML.
\item We use the type system of the metalanguage
    to check statically that an object program is well-typed and closed.
\item We show clean, comparable implementations in MetaOCaml and Haskell.
\item We specify a functor signature that encompasses all our
  interpreters, from evaluation and compilation (\S\ref{language}) 
   to partial evaluation%
% aplas: commented
%  and CPS transformation (\S\ref{PE}).
.
\item We point a clear way to extend the object language with more features
    such as state%
%aplas (\S\ref{state})%
%aplas
\footnote{again, see code}
.
\item We describe an approach to self\hyp interpretation compatible with the
  above%
%aplas: (\S\ref{selfinterp})
.  Self\hyp interpretation turned out
  harder than expected%
%aplas
\footnotemark[\value{footnote}]
.
\end{enumerate*}
Our code is surprisingly simple and obvious in hindsight, but
it has been an open problem to
interpret a typed object language in a typed metalanguage without
tagging or type\hyp system extensions.  For example, \citet{taha-tag}
say that ``expressing such an interpreter in a statically typed
programming language is a rather subtle matter. In fact, it is only
recently that some work on programming type-indexed values in ML
\cite{yang-encoding} has given a hint of how such a function can be
expressed.''  We discuss related work in~\S\ref{related}.

@@

\subsection{How to make encoding flexible: abstract the interpreter}
\label{encoding}
We embed our language in (Meta)OCaml and Haskell.  In Haskell,
the functions that construct object terms are methods in a type class
|Symantics| (with a parameter |repr| of kind |* -> *|)%
%aplas-specific
, Fig.~\ref{fig:symantics-haskell}(a)%
. The class is so named
because its interface gives the syntax of the object language and its
instances give the semantics.
\begin{figure}
%
(a) \begin{code2}
class Symantics repr where
  int  :: Int  -> repr Int       bool :: Bool -> repr Bool
  lam :: (repr a -> repr b) -> repr (a -> b)
  app :: repr (a -> b) -> repr a -> repr b
  fix :: (repr a -> repr a) -> repr a

>>>>

It is commonly thought that to interpret a typed object language in
a typed metalanguage while preserving types is difficult and requires
GADTs or dependent types \citep{taha-tag}.  In fact, this problem
motivated much work on GADTs \cite{xi-guarded,peyton-jones-simple} and
on dependent types \cite{WalidICFP02,fogarty-concoqtion}.
%aplas-else
\begin{comment}
For a metalanguage's type system to allow the well-typed object term
|test1| but disallow the ill-typed object term |test2|, fancier types
such as GADTs or dependent types seem necessary.  
\end{comment}
Yet other type systems
have been proposed to distinguish closed terms like |test1| from open
terms 
%aplas-comment: like |test3|
%aplas \cite{WalidPOPL03,NanevskiICFP02,NanevskiJFP05,DaviesJACM01,nanevski-contextual},
\cite{WalidPOPL03,NanevskiJFP05,DaviesJACM01},
so that |lookup| never receives an empty environment.  We discuss these
proposals further in \S\ref{related}%
%aplas-specific
.
%aplas-else
\begin{comment}
; here we just note that many
advanced type systems have been devised to ensure statically that an
object term is well-typed and closed.
\end{comment}

@@

In sum, by Church\hyp encoding terms using ordinary functions, we
achieve a tagless evaluator for a typed object language in a
metalanguage with a simple 
%aplas
Hindley-Milner type system%
%aplas \cite{hindley-principal,milner-theory}
.  In this \emph{final} rather
than \emph{initial} approach, both kinds of run-time errors in
\S\ref{tagproblem} (applying a nonfunction and evaluating an open
term) are reported at compile time. Because the new interpreter
uses no universal type or pattern matching, it never results in a
run-time error (and is in fact total).  Because this safety is obvious
not just to us but also to the metalanguage implementation, we avoid
the serious performance penalty \cite{WalidICFP02} of error checking.
\Citet{Gluck-jones-optimality} explains deeper technical reasons that
inevitably lead to these performance penalties.

@@

The term ``constructor'' functions |b|, |lam|, |app|, and so on appear
free in the encoding of an object term such as |testf1| above.  Defining
these functions differently gives rise to different interpreters, that
is, different folds on object programs.  Given the same term
representation but varying the interpreter, we can
\begin{itemize*}
    \item evaluate the term to a value in the metalanguage;
    \item measure the size or depth of the term;
    \item compile the term, with staging support such as in MetaOCaml;
    \item even partially evaluate the term, online; and
    \item transform the term to continuation\hyp passing style (CPS),
        including call-by-name (CBN) CPS, so to isolate the evaluation
        order of the object language from that of the metalanguage
%aplas
\footnote{Due to serious lack of space, 
we refer the reader to the accompanying code for this.}%
.
\end{itemize*}
We have programmed our interpreters in OCaml (and, for staging,
\cite{metaocaml}) and standard Haskell. The complete code is available
\url{http://okmij.org/ftp/packages/tagless-final.tar.gz}
to supplement the paper. Our examples below switch between (Meta)OCaml
and Haskell even though we have implemented each example equivalently in
both metalanguages, because some of our claims are more obvious in one
metalanguage than the other.  For example, MetaOCaml provides
convenient, typed staging facilities.

@@

\begin{enumerate*}
\item We build interpreters that evaluate (\S\ref{language}),
    compile (or evaluate with staging) (\S\ref{compiler}), and partially evaluate (\S\ref{PE}) a typed higher-order object language
   in a typed metalanguage, in direct and continuation\hyp passing styles%
% aplas
%   (\S\ref{variations}).
.
\item All these interpreters use no type tags, patently never get stuck,
    and need no advanced type-system features such as GADTs, dependent types,
    or intentional type analysis.
\item The partial evaluator avoids polymorphic lift and delays binding-time
    ana-lysis.  It bakes a type-to-type map into the interpreter
    interface to eliminate the need for GADTs and thus remain portable
    across Haskell 98 and ML.
\item We use the type system of the metalanguage
    to check statically that an object program is well-typed and closed.
\item We show clean, comparable implementations in MetaOCaml and Haskell.
\item We specify a functor signature that encompasses all our
  interpreters, from evaluation and compilation (\S\ref{language}) 
   to partial evaluation%
% aplas: commented
%  and CPS transformation (\S\ref{PE}).
.
\item We point a clear way to extend the object language with more features
    such as state%
%aplas (\S\ref{state})%
%aplas
\footnote{again, see code}
.
\item We describe an approach to self\hyp interpretation compatible with the
  above%
%aplas: (\S\ref{selfinterp})
.  Self\hyp interpretation turned out
  harder than expected%
%aplas
\footnotemark[\value{footnote}]
.
\end{enumerate*}
Our code is surprisingly simple and obvious in hindsight, but
it has been an open problem to
interpret a typed object language in a typed metalanguage without
tagging or type\hyp system extensions.  For example, \citet{taha-tag}
say that ``expressing such an interpreter in a statically typed
programming language is a rather subtle matter. In fact, it is only
recently that some work on programming type-indexed values in ML
\cite{yang-encoding} has given a hint of how such a function can be
expressed.''  We discuss related work in~\S\ref{related}.

@@

\subsection{How to make encoding flexible: abstract the interpreter}
\label{encoding}
We embed our language in (Meta)OCaml and Haskell.  In Haskell,
the functions that construct object terms are methods in a type class
|Symantics| (with a parameter |repr| of kind |* -> *|)%
%aplas-specific
, Fig.~\ref{fig:symantics-haskell}(a)%
. The class is so named
because its interface gives the syntax of the object language and its
instances give the semantics.
\begin{figure}
%
(a) \begin{code2}
class Symantics repr where
  int  :: Int  -> repr Int;       bool :: Bool -> repr Bool
  lam :: (repr a -> repr b) -> repr (a -> b)
  app :: repr (a -> b) -> repr a -> repr b
  fix :: (repr a -> repr a) -> repr a
